Presheaf Model for Dependent Types

### 2017-09-28 papers/talk-11/paper-11.tex
Rewrite section on the model

@@

\begin{definition}[Category of families of sets]
  The category of families of sets, $\fcat{Fam}$, has as objects $(A, B)$ where
  $A$ is a set and $B = (B_a~|~a \in A)$ is an $A$-indexed family of sets $B_a$,
  $a \in A$. A morphism $(A, B) \to (A', B')$ is given by a pair $(f, g)$ where
  $f: A \to A'$ and $g$ is an $A$-indexed famiy of functions such that $g_a: B_a
  \to B_{fa}'$.
\end{definition}

@@

  \item[$\bullet$] $F$ is a functor $F\from \cat{C}^\op \to \fcat{Fam}$; For a
    context $\Gamma$ we write $Ty_F(\Gamma)$ for the indexing set of the family
    $F(\Gamma)$ and its family as $(Ter_F(\Gamma;A)~|~A \in Ty_F(\Gamma))$;
    $\Gamma \vdash A~type$ means that $A \in Ty_F(\Gamma)$ and $\Gamma \vdash a
    : A$ means $a \in Ter_F(\Gamma;A)$. Note that for a context morphism
    $\sigma \from \Delta \to \Gamma$ the morphism $F(\sigma)$ acts on types
    $\Gamma \vdash A~type$ as $A\sigma$ and on terms $\Gamma \vdash a : A$ as
    $a\sigma$.
\end{itemize}
\end{definition}

@@

Let us now consider the categories of presheaves and see that they form a
category with families where the contexts are presheaves.  Let $\cat{C}$ be a
small category. We will refer to the objects of $\cat{C}$  as $I,J,K$ and
morphisms as $f,g,h$.

The context $\Gamma$ is then a presheaf on $\cat{C}$, $\Gamma \from \cat{C}^\op
\to \Set$, that is, we are given a set $\Gamma(I)$ for each $I$ in
$\ob(\cat{C})$, and functions $\Gamma(I) \to \Gamma(J)$, $\rho \mapsto \rho f$
(called \emph{restriction} and written as $f$ acting on the right) for each $f:
J \to I$ such that
\[
  \rho 1 = \rho \quad \text{and} \quad (\rho f) g = \rho (f g)
\]
for $g: K \to J$ and $f: J \to I$. We write $\Gamma_f$ for $\rho \mapsto \rho
f$.

The empty context $\diamond$ is a terminal presheaf and a context morphism
$\sigma \from \Delta \to \Gamma$ is a natural transformation that makes the
diagram commute.
\[
  \xymatrix{
    \Delta(I) \ar[r]^{\sigma_I} \ar[d]^{\Delta_f} & \Gamma(I) \ar[d]^{\Gamma_f} \\
    \Delta(J) \ar[r]^{\sigma_J} & \Gamma(J)
  }
\]

In the following, we will omit the indices on $\sigma$; Then if we write
$\Gamma_f$ and $\Delta_f$ as $f$ acting on the right (recall the session on
presheaves), the equation becomes:
\[
  (\sigma \rho) f = \sigma (\rho f)
\]
for $\rho$ in $\Delta(I)$.

Now we consider how to give a dependent type $\Gamma \vdash A~type$ in context
$\Gamma$. For each object $I \in \ob(\cat{C})$ and $\rho \in \Gamma(I)$ we require a
set $A \rho$, and for each $f \from J \to I$, we require a function $A \rho \to
A (\rho f)$, written as $a \mapsto af$ satisfying $a1 = a$ and $afg = a (f g)$
if $g \from K \to J$. The dependence on $I$ in $A\rho$ and on $I$ and $\rho$ in
$af$ is omitted for brevity. Substitution $\Delta \vdash A \sigma$ with $\sigma
\from \Delta \to \Gamma$ is simply given $(A \sigma) \rho = A (\sigma \rho)$
for $\rho \in \Gamma(I)$, together with the induced map
\[
  (A\sigma)\rho = A (\sigma \rho) \to A ((\sigma \rho)f) = A (\sigma (\rho f)) = (A \sigma) (\rho f)
\]
for $f \from J \to I$.

The definition of a dependent type can be rephrased more categorically: $A$ is
a presheaf on $\int_C \Gamma$, where $\int_C \Gamma$ is a category of elements
of the presheaf $\Gamma$, defined as follows:
\begin{itemize}
  \item objects are pairs $(I, \rho)$, where $I \in \ob(\cat{C})$ and $\rho \in
    \Gamma(I)$
  \item a morphism $(J, \rho') \to (I, \rho)$ is a morphism $f \from J \to I$
    in $\cat{C}$ such that $\rho' = \Gamma_f \rho$
\end{itemize}

A term $\Gamma \vdash a : A$ of a dependent type $\Gamma \vdash A~type$ is
given by a family of elements $a \rho \in A \rho$ for each $I \in \ob(\cat{C})$
and $\rho \in \Gamma(I)$, such that $a \rho f = a (\rho f)$ for each $f \from J
\to I$. The substitution $\Delta \vdash a \sigma : A \sigma$ with $\sigma \from
\Delta \to \Gamma$ is given by the family $(a \sigma) \rho = a (\sigma \rho)$.

For $\Gamma \vdash A~type$ the context extension $\vdash \Gamma.A~ctxt$ is
defined by
\begin{align*}
  &(\Gamma.A)(I) = \{(\rho, u)~|~\rho \in \Gamma(I) \land u \in A \rho\} &&
  \text{for $I \in \cat{C}$}\\
  &(\rho, u)f = (\rho f, u f) && \text{for $f \from J \to I$}
\end{align*}
and the projections are defined by $\mathbf{p} \from \Gamma.A \to \Gamma$,
$\mathbf{p}(\rho, u) = \rho$, and $\Gamma.A \vdash \mathbf{q} : Ap$ by
$\mathbf{q}(\rho, u) = u$.  Now assume $\sigma \from \Delta \to \Gamma$ and
$\Delta \vdash a : A\sigma$; we define $(\sigma, a) \from \Delta \to \Gamma.A$
by $(\sigma, a)\rho = (\sigma \rho, a \rho)$.

@@

\begin{example}
  The type for the function $A \to A$ (where $A$ is a ground type) is expressed
  as an exponential $M(A)^{M(A)}$ in STLC with $M$ being the interpretation
  function and $M(A)$ an object corresponding to the ground type in cartesian
  closed category.
\end{example}

>>>>

\newcommand{\Fam}{\fcat{Fam}}

@@

\begin{definition}[Category of families of sets]
  The category of families of sets, $\Fam$, has as objects $(A, B)$ where
  $A$ is a set and $B = (B_a~|~a \in A)$ is an $A$-indexed family of sets $B_a$,
  $a \in A$. A morphism $(A, B) \to (A', B')$ is given by a pair $(f, g)$ where
  $f: A \to A'$ and $g$ is an $A$-indexed famiy of functions such that $g_a: B_a
  \to B_{fa}'$.
\end{definition}

@@

  \item[$\bullet$] $F$ is a functor $F\from \cat{C}^\op \to \Fam$; For a
    context $\Gamma$ we write $Ty_F(\Gamma)$ for the indexing set of the family
    $F(\Gamma)$ and its family as $(Ter_F(\Gamma;A)~|~A \in Ty_F(\Gamma))$;
    $\Gamma \vdash A~type$ means that $A \in Ty_F(\Gamma)$ and $\Gamma \vdash a
    : A$ means $a \in Ter_F(\Gamma;A)$. Note that for a context morphism
    $\sigma \from \Delta \to \Gamma$ the morphism $F(\sigma)$ acts on types
    $\Gamma \vdash A~type$ as $A\sigma$ and on terms $\Gamma \vdash a : A$ as
    $a\sigma$.

  \item[$\bullet$] There are operations of context extension as defined above.
\end{itemize}
\end{definition}

@@

In the following we will give a presheaf model for dependent types. First
recall the definition of a presheaf:

\begin{definition}[Presheaf]
  Let $\cat{C}$ be a category. A \emph{presheaf} on $\cat{C}$ is a functor
  $\cat{C}^\op \to \Set$.
\end{definition}

Let us see how the category of presheaves over $\cat{C}$ (we denote it as
$Psh(\cat{C})$) forms a category of families where presheaves are contexts.  If
we set presheaves to be the contexts, then the context morphisms are morphisms
in $[\cat{C}^\op, \Set]$, which are natural transformations between contravariant
functors. These natural transformations satisfy the following naturality
condition:
\[
  \xymatrix{
    \Delta(I) \ar[r]^{\sigma_I} \ar[d]^{\Delta_f} & \Gamma(I) \ar[d]^{\Gamma_f} \\
    \Delta(J) \ar[r]^{\sigma_J} & \Gamma(J)
  }
\]
where $I, J \in \ob(\cat{C}); \Gamma, \Delta \in Psh(\cat{C}); \sigma
\from \Delta \to \Gamma$, and $f \from J \to I$. $\Gamma_f$ and $\Delta_f$
denote the morphisms that we obtain from the morphism $f$ in $\Set$ by the
functoriality of $\Gamma$ and $\Delta$. For instance, if $\rho \in \Gamma(I)$,
then we have a morphism $\Gamma_f$ to the corresponding $\rho f \in \Gamma(J)$
(we write $f$ applied on the right to connect to the initial morphism in
$\cat{C}$) since the contravariant functor $\Gamma$ preserves morphism
mappings.

Recall that every context has a morphism from itself to the empty context.
Hence we set the empty context to be the terminal object in the category of
presheaves, the terminal context, which is a functor that maps everything to
the terminal object in $\Set$, the singleton set. Indeed, we need a unique
natural transformation from every presheaf to the terminal presheaf. This
natural transformation is a family of unique morphisms to the terminal object
in $\Set$.

To define $F$ let us consider a category of elements of a presheaf $\Gamma$,
denoted by $\int\Gamma$. The objects of $\int \Gamma$ are $(I, \rho)$ where
$I \in \ob(\cat{C})$ and $\rho \in \Gamma(I)$. For $J, I \in \ob(\cat{C})$ the
morphism between $(J, \rho')$ and $(I, \rho)$ is a morphism $f \from J \to I$
in $\cat{C}$ denoted as $(f, \rho)$, such that $\rho' = \Gamma_f \rho$. Note
that this way the composition is inherited from $\cat{C}$.

Now we can define $Ty_F(\Gamma)$ as a set of presheaves over $\int \Gamma$.
Then $A \in Ty_F(\Gamma)$ is a presheaf and for a morphism in $\cat{C}$, $f
\from J \to I$, $\rho \in \Gamma(I)$, $x \in A(I, \rho)$ we have $A(f, \rho)(x)
\in A(J, \Gamma_f \rho)$ since $A$ is a functor. For a context morphism $\sigma
\from \Delta \to \Gamma$ we define a functor $\int \sigma \from \int \Delta \to
\int \Gamma$ by
\[
 \int \sigma (I, \rho') = (I, \sigma_I \rho'),
\]
where $\rho' \in \Delta(I)$.

To define terms of a dependent type $\Gamma \vdash A~type$ we require a family
of elements $a \rho \in A \rho$ for each $I \in \cat{C}$ and $\rho \in
\Gamma(I)$. These elements should satisfy for any $f \from J \to I$ the
equation
\[
  A(f, \rho)(a(I, \rho)) = a(J, \Gamma_f \rho).
\]
The context morphism $\sigma \from \Delta \to \Gamma$ then operates on a term
$\Delta \vdash a\sigma : A\sigma$ by composition
\[
  (a\sigma_I)(I, \rho) = a(I, \sigma_I(\rho)).
\]
We define the presheaf $\Gamma.A$ corresponding to a context extension as
\[
  (\Gamma.A)(I) = \{(\rho, u)~|~\rho \in \Gamma(I) \land u \in A(I, \rho)\}
\]
for objects $I \in \cat{C}$ and for morphisms $f \from J \to I$ as
\[
  (\Gamma.A)_f(\rho, u) = ((\Gamma.A)_f \rho, A(f, \rho)(u)).
\]
The projections are then defined by $\mathbf{p} \from \Gamma.A \to \Gamma$,
$\mathbf{p}(\rho, u) = \rho$, and $\Gamma.A \vdash \mathbf{q} : A\mathbf{p}$ by
$\mathbf{q}(\rho, u) = u$. Under context morphism $\sigma \from \Delta \to
\Gamma$ we define an extension of $\Gamma$ with a term $\Delta \vdash a :
A\sigma$ $(\sigma, a) \from \Delta \to \Gamma.A$ by $(\sigma, a)\rho = (\sigma
\rho, a \rho)$.

@@

So far we know how to represent a type $\Gamma \vdash A~type$ and a dependent
type $\Gamma.A \vdash B$. Now we want to have a representation for a function
that corresponds to a type $\Gamma \vdash \Pi A. B$.
